---
title: "Prepare dashboard data"
author: "Jasmine Lee"
date: "2026-07-26"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
editor_options:
  markdown:
    wrap: 72
---

# 0. OVERVIEW & SETUP

### **Setup**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

options(repos = c(CRAN = "https://cloud.r-project.org"))

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, countrycode)

# Read in cleaned dataset (not committed to the repository)
data <- readRDS("data.rds")

data %>% nrow()
data %>% dplyr::select(country, religion_wrd, life_satisfaction, happiness,
                       health, age, sex, survey_weight) %>% glimpse()
```

Helper functions

```{r helper-functions}
w_mean <- function(x, w) {
  keep <- !is.na(x) & !is.na(w) & w > 0
  if (sum(keep) == 0) return(NA_real_)
  sum(x[keep] * w[keep]) / sum(w[keep])
}

w_se <- function(x, w) {
  keep <- !is.na(x) & !is.na(w) & w > 0
  if (sum(keep) < 2) return(NA_real_)
  x <- x[keep]; w <- w[keep]
  n_eff <- sum(w)^2 / sum(w^2)          # Kish effective sample size
  m     <- sum(w * x) / sum(w)
  v     <- sum(w * (x - m)^2) / sum(w)  # weighted variance
  sqrt(v / n_eff)
}
```

# 1. CATEGORISE MAJORITY/MINORITY STATUS 

Individuals will be classified as belonging to majority or minority
religious groups according to the World Values Survey overall sample.

## Clean outcomes & weights

```{r clean-outcomes}
data <- data %>%
  mutate(
    life_satisfaction = if_else(life_satisfaction < 0, NA_real_, as.numeric(life_satisfaction)),
    happiness         = as.numeric(happiness),
    health            = as.numeric(health),
    survey_weight     = if_else(is.na(survey_weight) | survey_weight <= 0, 1, survey_weight)
  )

# Keep only respondents with a religious affiliation reported
data <- data %>% filter(!is.na(religion_wrd))

data %>% nrow()
```

## Country religious composition (according to WVS sample)

```{r wvs-composition}
wvs_composition <- data %>%
  group_by(country) %>%
  mutate(country_w = sum(survey_weight)) %>%
  group_by(country, religion_wrd) %>%
  summarise(
    n         = n(),
    n_w       = sum(survey_weight),
    country_w = first(country_w),
    .groups   = "drop"
  ) %>%
  mutate(share = 100 * n_w / country_w) %>%
  arrange(country, desc(share))

wvs_composition %>% slice_head(n = 10)
```

## Identify the majority group

Majority definition: the religious group holding **more than 50%** of
the weighted WVS sample in that country. Countries with no such group
are dropped.

```{r identify-majority}
country_majority <- wvs_composition %>%
  group_by(country) %>%
  slice_max(share, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(
    country,
    majority_religion = religion_wrd,
    majority_share    = share,
    has_majority      = share > 50
  )

# Countries with no majority group in the WVS sample
country_majority %>% filter(!has_majority) %>% arrange(desc(majority_share))

country_majority <- country_majority %>% filter(has_majority)

data <- data %>%
  inner_join(country_majority %>% dplyr::select(country, majority_religion),
             by = "country")

data$country %>% n_distinct()
```

## Categorise individuals

```{r categorise-individuals}
data <- data %>%
  mutate(
    religion_status = if_else(religion_wrd == majority_religion, "Majority", "Minority"),
    religion_status = factor(religion_status, levels = c("Majority", "Minority"))
  )

data %>% count(religion_status)
```

## Drop countries with very small minority samples

Countries where the minority makes up less than 1% of the sample are
dropped — the comparison would not be meaningful.

```{r drop-small-minorities}
country_check <- data %>%
  group_by(country) %>%
  summarise(
    n_total     = n(),
    n_majority  = sum(religion_status == "Majority"),
    n_minority  = sum(religion_status == "Minority"),
    pct_minority = 100 * n_minority / n_total,
    .groups = "drop"
  ) %>%
  arrange(pct_minority)

country_check %>% print(n = Inf)

countries_drop <- country_check %>%
  filter(pct_minority < 1 | n_minority < 30) %>%
  pull(country)

countries_drop

data <- data %>% filter(!country %in% countries_drop)

data$country %>% n_distinct()
data %>% nrow()
```
