---
title: "Prepare dashboard data"
author: "Jasmine Lee"
date: "2026-07-26"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
editor_options:
  markdown:
    wrap: 72
---

# 0. OVERVIEW & SETUP

### **Setup**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

options(repos = c(CRAN = "https://cloud.r-project.org"))

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, countrycode)

# Read in cleaned dataset (not committed to the repository)
data <- readRDS("data.rds")

data %>% nrow()
data %>% dplyr::select(country, religion_wrd, life_satisfaction, happiness,
                       health, age, sex, survey_weight) %>% glimpse()
```

Helper functions

```{r helper-functions}
w_mean <- function(x, w) {
  keep <- !is.na(x) & !is.na(w) & w > 0
  if (sum(keep) == 0) return(NA_real_)
  sum(x[keep] * w[keep]) / sum(w[keep])
}

w_se <- function(x, w) {
  keep <- !is.na(x) & !is.na(w) & w > 0
  if (sum(keep) < 2) return(NA_real_)
  x <- x[keep]; w <- w[keep]
  n_eff <- sum(w)^2 / sum(w^2)          # Kish effective sample size
  m     <- sum(w * x) / sum(w)
  v     <- sum(w * (x - m)^2) / sum(w)  # weighted variance
  sqrt(v / n_eff)
}
```

# 1. CATEGORISE MAJORITY/MINORITY STATUS

Individuals will be classified as belonging to majority or minority
religious groups according to the World Values Survey overall sample.

## Clean outcomes & weights

```{r clean-outcomes}
data <- data %>%
  mutate(
    life_satisfaction = if_else(life_satisfaction < 0, NA_real_, as.numeric(life_satisfaction)),
    happiness         = as.numeric(happiness),
    health            = as.numeric(health),
    survey_weight     = if_else(is.na(survey_weight) | survey_weight <= 0, 1, survey_weight)
  )

# Keep only respondents with a religious affiliation reported
data <- data %>% filter(!is.na(religion_wrd))

data %>% nrow()
```

## Country religious composition (according to WVS sample)

```{r wvs-composition}
wvs_composition <- data %>%
  group_by(country) %>%
  mutate(country_w = sum(survey_weight)) %>%
  group_by(country, religion_wrd) %>%
  summarise(
    n         = n(),
    n_w       = sum(survey_weight),
    country_w = first(country_w),
    .groups   = "drop"
  ) %>%
  mutate(share = 100 * n_w / country_w) %>%
  arrange(country, desc(share))

wvs_composition %>% slice_head(n = 10)

tidy_religion <- function(x) {
  x %>%
    gsub("_", " ", .) %>%
    gsub("No denomination", "No religious denomination", .) %>%
    gsub("^Jew$", "Jewish", .) %>%
    gsub("Ethnic religionist", "Ethnic religions", .) %>%
    gsub("Chinese folk religionist", "Chinese folk religions", .) %>%
    gsub("Other religion$", "Other", .)
}
```

## Identify the majority group

Majority definition: the religious group holding **more than 50%** of
the weighted WVS sample in that country. Countries with no such group
are dropped.

```{r identify-majority}
country_majority <- wvs_composition %>%
  group_by(country) %>%
  slice_max(share, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(
    country,
    majority_religion = religion_wrd,
    majority_share    = share,
    has_majority      = share > 50
  )

# Countries with no majority group in the WVS sample
country_majority %>% filter(!has_majority) %>% arrange(desc(majority_share))

country_majority <- country_majority %>% filter(has_majority)

data <- data %>%
  inner_join(country_majority %>% dplyr::select(country, majority_religion),
             by = "country")

data$country %>% n_distinct()
```

## Categorise individuals

```{r categorise-individuals}
data <- data %>%
  mutate(
    religion_status = if_else(religion_wrd == majority_religion, "Majority", "Minority"),
    religion_status = factor(religion_status, levels = c("Majority", "Minority"))
  )

data %>% count(religion_status)
```

## Drop countries with very small minority samples

Countries where the minority makes up less than 1% of the sample are
dropped — the comparison would not be meaningful.

```{r drop-small-minorities}
country_check <- data %>%
  group_by(country) %>%
  summarise(
    n_total     = n(),
    n_majority  = sum(religion_status == "Majority"),
    n_minority  = sum(religion_status == "Minority"),
    pct_minority = 100 * n_minority / n_total,
    .groups = "drop"
  ) %>%
  arrange(pct_minority)

country_check %>% print(n = Inf)

countries_drop <- country_check %>%
  filter(pct_minority < 1 | n_minority < 30) %>%
  pull(country)

countries_drop

data <- data %>% filter(!country %in% countries_drop)

data$country %>% n_distinct()
data %>% nrow()
```

# 2. BUILD DASHBOARD TABLES

## Country lookup

```{r country-lookup}
country_lookup <- data %>%
  distinct(country) %>%
  mutate(iso3c = countrycode(country, origin = "country.name", destination = "iso3c"))

# Check nothing failed to match
country_lookup %>% filter(is.na(iso3c))
```

## Table 1: Religious composition by country

```{r table-composition}
tbl_composition <- wvs_composition %>%
  semi_join(country_lookup, by = "country") %>%
  left_join(country_majority %>% dplyr::select(country, majority_religion), by = "country") %>%
  mutate(status = if_else(religion_wrd == majority_religion, "Majority", "Minority")) %>%
  transmute(country, religion = religion_wrd, n, share, status) %>%
  arrange(country, desc(share))

tbl_composition %>% slice_head(n = 10)

tbl_composition <- tbl_composition %>% mutate(religion = tidy_religion(religion))
tbl_gaps <- tbl_gaps %>% mutate(majority_religion = tidy_religion(majority_religion))
```

## Table 2: Outcome means by country and religious minority/majority status (means)

```{r table-means}
outcome_labels <- c(
  life_satisfaction = "Life satisfaction (1–10)",
  happiness         = "Happiness (1–4)",
  health            = "Self-rated health (1–5)"
)

tbl_means <- data %>%
  dplyr::select(country, religion_status, survey_weight,
                life_satisfaction, happiness, health) %>%
  pivot_longer(c(life_satisfaction, happiness, health),
               names_to = "outcome", values_to = "value") %>%
  filter(!is.na(value)) %>%
  group_by(country, outcome, religion_status) %>%
  summarise(
    n    = n(),
    mean = w_mean(value, survey_weight),
    se   = w_se(value, survey_weight),
    .groups = "drop"
  ) %>%
  mutate(
    lower = mean - 1.96 * se,
    upper = mean + 1.96 * se,
    outcome_label = outcome_labels[outcome]
  )

tbl_means %>% slice_head(n = 6)
```

## Table 3: Minority–majority gaps

Negative values indicate that religious minorities score lower than the
majority in that country.

```{r table-gaps}
tbl_gaps <- tbl_means %>%
  dplyr::select(country, outcome, outcome_label, religion_status, mean, se, n) %>%
  pivot_wider(
    names_from  = religion_status,
    values_from = c(mean, se, n)
  ) %>%
  mutate(
    gap    = mean_Minority - mean_Majority,
    gap_se = sqrt(se_Minority^2 + se_Majority^2),
    gap_lower = gap - 1.96 * gap_se,
    gap_upper = gap + 1.96 * gap_se
  ) %>%
  left_join(country_lookup, by = "country") %>%
  left_join(
    country_majority %>% dplyr::select(country, majority_religion, majority_share),
    by = "country"
  ) %>%
  left_join(
    country_check %>% dplyr::select(country, n_total, pct_minority),
    by = "country"
  )

tbl_gaps %>%
  filter(outcome == "life_satisfaction") %>%
  arrange(gap) %>%
  dplyr::select(country, majority_religion, gap, gap_lower, gap_upper) %>%
  print(n = Inf)
```

## Table 4: Missingness by status

Important note: minorities are less likely to report happiness outcomes
which means that the descriptive gaps are likely conservative.

```{r table-missingness}
tbl_missing <- data %>%
  group_by(religion_status) %>%
  summarise(
    n = n(),
    pct_missing_ls     = round(100 * mean(is.na(life_satisfaction)), 1),
    pct_missing_happy  = round(100 * mean(is.na(happiness)), 1),
    pct_missing_health = round(100 * mean(is.na(health)), 1),
    .groups = "drop"
  )

tbl_missing
```

# 3. SAVE AGGREGATED DATA FOR THE DASHBOARD

```{r save-dashboard-data}
dashboard_data <- list(
  composition = tbl_composition,
  means       = tbl_means,
  gaps        = tbl_gaps,
  missingness = tbl_missing,
  lookup      = country_lookup,
  meta = list(
    n_countries   = n_distinct(data$country),
    n_respondents = nrow(data),
    n_majority    = sum(data$religion_status == "Majority"),
    n_minority    = sum(data$religion_status == "Minority"),
    built_on      = Sys.Date()
  )
)

dir.create("data", showWarnings = FALSE)
saveRDS(dashboard_data, "data/dashboard_data.rds")

# Also export a plain CSV of the country-level gaps for download in the app
write_csv(dashboard_data$gaps, "data/country_gaps.csv")

dashboard_data$meta
```
